---
title: "p8105_hw2_msj2167"
output: github_document
date: "2026-10-07"
---

Loading libraries 

```{r message = F}
library(tidyverse)
library(readxl)
```

Importing Mr.Trash Wheel dataset 

Cleaned all the datasets to be able to bind together by making all variables in the same format. 

```{r}
mr_trash_wheel_df =
  read_excel("Data/202610 Trash Wheel Collection Data.xlsx", 
  sheet = "Mr. Trash Wheel",
  range = "A2:N742",
  na=c(".", "", "NA")) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
    mutate(
      sports_balls= as.integer(round(sports_balls)),
      trash_wheel = "Mr. Trash Wheel",
      year = as.numeric(year),
    )
```

Import Professor Trash Wheel

```{r warning = FALSE}
prof_trash_wheel_df =
  read_excel("Data/202610 Trash Wheel Collection Data.xlsx", 
  sheet = "Professor Trash Wheel",
  range = "A2:M139",
  na=c(".", "", "NA")) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
    mutate(
   trash_wheel = "Professor Trash Wheel",
 plastic_bottles = as.numeric(plastic_bottles)
 )
```

Import Gwynnda

```{r}
gwynnda_df =
  read_excel("Data/202610 Trash Wheel Collection Data.xlsx", 
  sheet = "Gwynnda the Good Wheel of the W",
  range = "A2:L394",
  na=c(".", "", "NA")) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
   mutate(
   trash_wheel = "Gwynnda")
```

Merging the datasets

```{r}
trash_df = 
  bind_rows(mr_trash_wheel_df, prof_trash_wheel_df, gwynnda_df)
```

In the trash_df dataset there are `r nrow(trash_df)` observations and `r ncol(trash_df)` variables. Some of the key variables are `r names(trash_df)`.

The total weight collected by Professor Trash Wheel was `r round(sum(pull(filter(trash_df, trash_wheel == "Professor Trash Wheel"), weight_tons), na.rm = TRUE), 2)` tons.

The total number of cigarette butts collected by Gwynnda in June of 2022 was `r format(sum(pull(filter(trash_df, trash_wheel == "Gwynnda", month == "June", year == 2022), cigarette_butts), na.rm = TRUE), big.mark = ",")`



