---
title: "p8105_hw2_msj2167"
output: github_document
date: "2026-10-07"
---

Loading libraries 

```{r message = F}
library(tidyverse)
library(readxl)
```

## Problem 2

Importing Mr.Trash Wheel dataset 

Cleaned all the datasets to be able to bind together by making all variables in the same format. 

```{r}
mr_trash_wheel_df =
  read_excel("Data/202610 Trash Wheel Collection Data.xlsx", 
  sheet = "Mr. Trash Wheel",
  range = "A2:N742",
  na=c(".", "", "NA")) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
    mutate(
      sports_balls= as.integer(round(sports_balls)),
      trash_wheel = "Mr. Trash Wheel",
      year = as.numeric(year),
    )
```

Import Professor Trash Wheel

```{r warning = FALSE}
prof_trash_wheel_df =
  read_excel("Data/202610 Trash Wheel Collection Data.xlsx", 
  sheet = "Professor Trash Wheel",
  range = "A2:M139",
  na=c(".", "", "NA")) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
    mutate(
   trash_wheel = "Professor Trash Wheel",
 plastic_bottles = as.numeric(plastic_bottles)
 )
```

Import Gwynnda

```{r}
gwynnda_df =
  read_excel("Data/202610 Trash Wheel Collection Data.xlsx", 
  sheet = "Gwynnda the Good Wheel of the W",
  range = "A2:L394",
  na=c(".", "", "NA")) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
   mutate(
   trash_wheel = "Gwynnda")
```

Merging the datasets

```{r}
trash_df = 
  bind_rows(mr_trash_wheel_df, prof_trash_wheel_df, gwynnda_df)
```

In the trash_df dataset there are `r nrow(trash_df)` observations and `r ncol(trash_df)` variables. Some of the key variables are `r names(trash_df)`.

The total weight collected by Professor Trash Wheel was `r round(sum(pull(filter(trash_df, trash_wheel == "Professor Trash Wheel"), weight_tons), na.rm = TRUE), 2)` tons.

The total number of cigarette butts collected by Gwynnda in June of 2022 was `r format(sum(pull(filter(trash_df, trash_wheel == "Gwynnda", month == "June", year == 2022), cigarette_butts), na.rm = TRUE), big.mark = ",")`

## Problem 3

Importing MCI_baseline dataset 

Ensure that sex and APOE4 carrier status are appropriate encoded (i.e. not numeric), and remove any participants who do not meet the stated inclusion criteria (i.e. no MCI at baseline). Discuss important steps in the import process and relevant features of the dataset. How many participants were recruited, and of these how many develop MCI? What is the average baseline age? What proportion of women in the study are APOE4 carriers?

When importing the data I created two dataframes, one to clean the data and that contained all participants who were recruited and another that filtered participants based on the inclusion criteria. When removing any participants who do not meet the stated inclusion criteria (MCI at baseline) I had to make sure to look into those who had NA for age_at_onset as it was missing if a subject remains MCI free during the follow-up period. However, I also had to look into those who had an age of onset greater than their current age at baseline, as they meet the inclusion criteria of being MCI free at baseline. 

```{r}
mci_recruitment_df =
  read_csv(
    file = "data/MCI_baseline.csv",
    skip = 1,
    na = c("", "NA", ".")) |>
  janitor::clean_names() |>
  mutate(
  sex = recode_values(
    sex,
    1 ~ "male",
    0 ~ "female"),
  apoe4 = recode_values(
    apoe4,
    1 ~ "carrier",
    0 ~ "non-carrier"))

mci_baseline_df =
  filter(mci_recruitment_df, is.na(age_at_onset) | age_at_onset > current_age)

```

There were `r nrow(mci_recruitment_df)` participants recruited. After excluding those with MCI at baseline, `r nrow(mci_baseline_df)` participants remained, of whom `r nrow(filter(mci_baseline_df, !is.na(age_at_onset)))` developed MCI during follow-up.

The average baseline age was `r round(mean(pull(mci_baseline_df, current_age), na.rm = TRUE), 1)`

The proportion of women who were APOE4 carriers `r round(mean(pull(filter(mci_baseline_df, sex == "female"), apoe4) == "carrier") * 100, 1)`%

Importing and cleaning longitudinally observed biomarker dataset

```{r}
mci_amyloid_df =
  read_csv(
    file = "data/mci_amyloid.csv",
    skip = 1,
    na = c("", "NA", ".")) |>
  janitor::clean_names()
```

